```{r setup, include=FALSE}
#  bookdown::render_book("index.Rmd", "bookdown::gitbook", output_dir = "docs/")

# packages
library(bookdown)
library(RColorBrewer)
library(readr)
library(plotly)
library(tidyverse)
library(reshape2)
library(latex2exp)
library(svglite)
library(knitr)
library(kableExtra)
library(formattable)
library(DT)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE, comment = "") #, dev = "svglite", fig.width=8
#plotly::config(plot_ly(), displaylogo = FALSE, mathjax = "cdn")
```

\newcommand{\Var}{\operatorname{Var}}
\newcommand{\E}{\operatorname{E}}
\newcommand{\se}{\mathsf{se}}

# (PART\*) Course Notes {-}

# Preliminaries {-}

## Notation {-}

Uppercase roman letters, e.g., $X$, will typically denote random variables (rvs); lower case letters, e.g., $x$, will denote a particular value (observation) of a rv. Rvs have probability distributions. Distributions are typically characterized by *parameters* which are fixed real numbers. Parameters describe population characteristics that are often unknown and must be estimated from data. Statistical inference is a tool that will help us to do this.

> ⚠️  Statistical models comprise both rvs and parameters. Be careful not to confuse them! 

## Abbreviations {-}

```{r notation-abbrev, echo=FALSE, warning = FALSE, message = FALSE}
datatable(tibble(read_csv("data/notation-abbrev.csv")), rownames = FALSE, #filter = 'bottom',
          options = list(pageLength = 5, scrollX = T), autoHideNavigation = TRUE,
          caption = "Collection of abbreviations that are used in these notes.")
# kable(tibble(read_csv("data/notation-abbrev.csv")))
```
