{{< include preamble.qmd >}}

# Sampling distributions {#sec-sampling-distributions}

A *statistic* is a quantity that can be calculated from sample data. Before observing data, a statistic is an unknown quantity and is, therefore, a rv. 

::: {#def-statistic}
## Statistic 

Let $X_1, \dots, X_n$ be observable rvs and let $g$ be an arbitrary real-valued function of $n$ random variables. The rv $$T = g(X_1, \dots, X_n)$$ is a statistic. 
:::

We refer to the probability distribution for a statistic as a sampling distribution. The sampling distribution illustrates how the statistic will vary across possible sample data. The sampling distribution contains information about the values a statistic is likely to assume and how likely it is to assume those values prior to observing data. 

::: {#def-sampling-dist}
## Sampling distribution

Suppose rvs $X_1, \dots, X_n$ are a random sample from $F(\theta),$ a distribution depending on a parameter $\theta$ whose value is unknown. Let the rv $$T = g(X_1, \dots, X_n, \theta)$$ be a function of $X_1, \dots, X_n$ and (possibly) $\theta$. The distribution of $T$ (given $\theta$) is the sampling distribution of $T$. 
:::

The sampling distribution of $T$ is derived from the distribution of the random sample. Often we will be interested in a statistic $T$ that is an estimator for a parameter $\theta$ (that is, $T$ will not depend on $\theta$). 

In what follows, we review several special families of distributions that are widely used in probability and statistics. These special families of distributions will be indexed by one or more parameters and include discrete distributions (Bernoulli, Binomial, Poisson, and discrete Uniform) as well as continuous distributions (continuous Uniform, Normal, Student's $\mathsf{t}$, $\chi^2$, and $\mathsf{F}$).


## Bernoulli distribution {#sec-bernoulli-distribution}

The Bernoulli distribution describes a single trial with two possible outcomes, often coded as $1$ (success) and $0$ (failure). It is the basic building block behind proportions and binomial counts.

::: {#def-bernoulli-dist}
## Bernoulli distribution

A discrete rv $X$ has a Bernoulli distribution with parameter $p \in (0,1)$ if
$$
P(X=x; p) =
\begin{cases}
 p, & x=1,\\
 1-p, & x=0,\\
 0, & \text{otherwise}.
\end{cases}
$$
Equivalently,
$$
P(X=x; p) = p^x(1-p)^{1-x}, \quad x\in\{0,1\}.
$$ {#eq-bernoulli-pmf}
We write $X \sim \mathsf{Bernoulli}(p)$.
:::

:::{.callout-warning}
## Parameter

The parameter $p$ is the probability of success, i.e. $p = P(X=1)$.
:::

If $X \sim \mathsf{Bernoulli}(p)$, then it can be shown that
$$
\E[X] = p
\qquad \text{and} \qquad
\Var(X) = p(1-p).
$$ {#eq-bernoulli-mean-var}

The pmf for two Bernoulli rvs with different values of $p$ is shown in @fig-bernoulli-pmf.

```{r}
#| label: fig-bernoulli-pmf
#| fig-cap: "The pmf of a Bernoulli rv for two values of the success probability $p$."
#| echo: false
#| warning: false
#| message: false
bernoulli_df <- expand.grid(x = c(0,1), p = c(0.2, 0.7)) |>
  mutate(prob = ifelse(x == 1, p, 1 - p),
         p_strip = paste0("p==", p))

ggplot(bernoulli_df, aes(x = x, y = prob)) +
  geom_col(alpha = 0.6) +
  facet_wrap(~ p_strip, nrow = 1, labeller = label_parsed) +
  labs(x = TeX("$x$"), y = TeX("$P(X=x)$")) +
  scale_x_continuous(breaks = c(0,1)) +
  theme_ur
```

::: {#exm-bernoulli}
Suppose that a component passes a quality check with probability $p = 0.95$. Let $X=1$ if the component passes and $X=0$ otherwise. Then $X \sim \mathsf{Bernoulli}(0.95)$, and the probability that the component fails is
$$
P(X=0) = 1 - p = 0.05.
$$
:::

:::{.callout-important}
## Indicator variables

A Bernoulli rv is often used as an **indicator**: it takes value $1$ when a property holds, and $0$ when it does not. This is exactly the setup used for estimating proportions in @sec-estimating-proportions.
:::


## Binomial distribution {#sec-binomial-distribution}

The binomial distribution describes the number of successes in $k$ independent Bernoulli trials, each having the same success probability $p$.

::: {#def-binomial-dist}
## Binomial distribution

Let $X$ denote the number of successes in $k$ independent trials, where each trial is a $\mathsf{Bernoulli}(p)$ rv. Then $X$ has a binomial distribution with parameters $k \in \mathbf{N}_{>}$ and $p \in (0,1)$ if
$$
P(X=x; k,p) = \binom{k}{x} p^x (1-p)^{k-x}, \quad x = 0,1,\dots,k.
$$ {#eq-binomial-pmf}
We write $X \sim \mathsf{Binomial}(k,p)$.
:::

:::{.callout-warning}
## Parameters

The binomial distribution has **two parameters**: the number of trials $k$ and the success probability $p$.
:::

If $X \sim \mathsf{Binomial}(k,p)$, then
$$
\E[X] = kp
\qquad \text{and} \qquad
\Var(X) = kp(1-p).
$$ {#eq-binomial-mean-var}

The pmf for $\mathsf{Binomial}(10,p)$ for several values of $p$ is shown in @fig-binomial-pmf.

```{r}
#| label: fig-binomial-pmf
#| fig-cap: "The pmf of $X \\sim \\mathsf{Binomial}(10,p)$ for several values of $p$."
#| echo: false
#| warning: false
#| message: false
k <- 10
p_vals <- c(0.2, 0.5, 0.8)

binom_df <- expand.grid(x = 0:k, p = p_vals) |>
  mutate(prob = dbinom(x, size = k, prob = p),
         p_strip = factor(paste0("p==", p),
                          levels = paste0("p==", p_vals)))

ggplot(binom_df, aes(x = x, y = prob)) +
  geom_col(alpha = 0.6) +
  facet_wrap(~ p_strip, nrow = 1, labeller = label_parsed) +
  labs(x = TeX("$x$"), y = TeX("$P(X=x)$")) +
  scale_x_continuous(breaks = 0:k) +
  theme_ur
```

::: {#exm-binomial}
A student answers $k=10$ multiple-choice questions by guessing. Each question has probability $p=0.25$ of being correct. If $X$ is the number of correct answers, then
$$
X \sim \mathsf{Binomial}(10, 0.25).
$$
The probability of getting exactly $3$ correct answers is
$$
P(X=3) = \binom{10}{3}(0.25)^3(0.75)^7.
$$
This can be computed in `R` as `dbinom(3, size = 10, prob = 0.25)`.
:::

:::{.callout-tip}
## Binomial counts are sums of Bernoulli trials

If $X_1,\dots,X_k \sim \mathsf{Bernoulli}(p)$ are independent, then
$$
X = \sum_{i=1}^k X_i \sim \mathsf{Binomial}(k,p).
$$
This is why binomial distributions appear naturally when counting the number of times an event occurs in repeated trials.
:::


## Poisson distribution {#sec-poisson-distribution}

The Poisson distribution is widely used to model the number of events in a fixed unit of exposure (often time or space), such as the number of arrivals at a desk per hour or the number of faults detected per kilometre.

::: {#def-poisson-dist}
## Poisson distribution

A discrete rv $X$ has a Poisson distribution with parameter $\lambda > 0$ if
$$
P(X=x; \lambda) = \frac{e^{-\lambda}\lambda^x}{x!}, \quad x = 0,1,2,\dots
$$ {#eq-poisson-pmf}
We write $X \sim \mathsf{Poisson}(\lambda)$.
:::

:::{.callout-warning}
## Parameter

The parameter $\lambda$ is the mean number of events in the unit of exposure (e.g. “per hour”).
:::

If $X \sim \mathsf{Poisson}(\lambda)$, then
$$
\E[X] = \lambda
\qquad \text{and} \qquad
\Var(X) = \lambda.
$$ {#eq-poisson-mean-var}

That is, the Poisson distribution has the distinctive property that the **mean equals the variance**. The pmf for several values of $\lambda$ is shown in @fig-poisson-pmf.

```{r}
#| label: fig-poisson-pmf
#| fig-cap: "The pmf of a Poisson rv for several values of the rate parameter $\\lambda$."
#| echo: false
#| warning: false
#| message: false
lambda_vals <- c(1, 4, 10)

poisson_df <- expand.grid(x = 0:25, lambda = lambda_vals) |>
  mutate(prob = dpois(x, lambda = lambda),
         lam_strip = factor(paste0("lambda==", lambda),
                            levels = paste0("lambda==", lambda_vals)))

ggplot(poisson_df, aes(x = x, y = prob)) +
  geom_col(alpha = 0.6) +
  facet_wrap(~ lam_strip, nrow = 1, labeller = label_parsed) +
  labs(x = TeX("$x$"), y = TeX("$P(X=x)$")) +
  scale_x_continuous(breaks = seq(0, 25, by = 5)) +
  theme_ur
```

::: {#exm-poisson}
Suppose the number of emails you receive in an hour is modelled as $X \sim \mathsf{Poisson}(\lambda)$ with $\lambda = 2.5.$ The probability of receiving no emails in the next hour is
$$
P(X=0) = e^{-2.5}\frac{2.5^0}{0!} = e^{-2.5}.
$$
In `R`, this is `dpois(0, lambda = 2.5)`.
:::

:::{.callout-note}
## When does Poisson modelling make sense?

Poisson models are commonly used when (i) events occur one at a time, (ii) they occur “at random” throughout the exposure period, and (iii) the average event rate is roughly constant over the period being observed.
:::

## Uniform Distribution {#sec-uniform-distribution}

The uniform distribution places equal weight on the items being sampled. The items can be discrete or be a continuum. 

### Discrete uniform distribution {#sec-discrete-uniform-distribution}

The discrete uniform places equal probability on a *finite* set of possible values.

::: {#def-discrete-uniform}
## Discrete uniform distribution

A discrete rv $X$ has a discrete uniform distribution on the integers $\{a,a+1,\dots,b\}$ with $a<b$, if
$$
P(X=x; a,b) =
\begin{cases}
\frac{1}{b-a+1}, & x \in \{a,a+1,\dots,b\},\\[4pt]
0, & \text{otherwise}.
\end{cases}
$$ {#eq-discrete-uniform-pmf}
We write $X \sim \mathsf{Unif}\{a,\dots,b\}$.
:::

:::{.callout-warning}
## Parameters

The parameters $a$ and $b$ determine the finite support $\{a,a+1,\dots,b\}$.
:::

If $X \sim \mathsf{Unif}\{a,\dots,b\}$, then
$$
\E[X] = \frac{a+b}{2}
\qquad \text{and} \qquad
\Var(X) = \frac{(b-a+1)^2 - 1}{12}.
$$ {#eq-discrete-uniform-mean-var}

A familiar example is the outcome of a fair six-sided die: $X \sim \mathsf{Unif}\{1,\dots,6\}$. The pmf is shown in @fig-discrete-unif-pmf.

```{r}
#| label: fig-discrete-unif-pmf
#| fig-cap: "The pmf of a discrete uniform rv on $\\{1,2,3,4,5,6\\}$ (a fair die)."
#| echo: false
#| warning: false
#| message: false
disc_unif_df <- data.frame(x = 1:6) |>
  mutate(prob = 1/6)

ggplot(disc_unif_df, aes(x = x, y = prob)) +
  geom_col(alpha = 0.6) +
  labs(x = TeX("$x$"), y = TeX("$P(X=x)$")) +
  scale_x_continuous(breaks = 1:6) +
  theme_ur
```

::: {#exm-discrete-uniform}
Let $X$ be the outcome of a fair die roll. Then $X \sim \mathsf{Unif}\{1,\dots,6\}.$ The probability of rolling a value at least $5$ is
$$
P(X \geq 5) = P(X=5) + P(X=6) = \frac{2}{6} = \frac{1}{3}.
$$
:::


### Continuous uniform distribution {#sec-continuous-uniform-distribution}

Compared with the discrete version, the continuous uniform distribution places equal weight across an interval.

::: {#def-uniform-dist}
## (Continuous) Uniform distribution

A continuous rv $X$ has a uniform distribution on $[a,b]$ with $a<b,$ if $X$ has pdf
$$
 f(x; a,b) = \frac{1}{b-a}\,, 
 \quad a < x < b\,,
$$
 or zero otherwise. 
We write $X \sim \mathsf{Unif}(a,b)$. 
:::  

::: {.callout-warning}
## Parameters

Note that $a$ and $b$ are parameters in @def-uniform-dist.
:::

::: {#exr-uniform}
As an exercise, derive the cdf using the definition. Derive a formula for the mean and variance in terms of the parameters $a$ and $b$.
:::

## Normal distribution {#sec-normal-distribution}

Normal distributions play an important role in probability and statistics as they describe many natural phenomena. For instance, the Central Limit Theorem tells us that the sample mean of a large random sample (size $m$) of rvs with mean $\mu$ and variance $\sigma^2$ is approximately normal in distribution with mean $\mu$ and variance $\sigma^2/m$.  

::: {#def-normal-dist}
## Normal or Gaussian distribution

A continuous rv $X$ has a normal distribution with parameters $\mu$ and $\sigma^2,$ where $-\infty < \mu < \infty$ and $\sigma > 0,$ if $X$ has pdf
$$
 f(x; \mu, \sigma) = \frac{1}{\sqrt{2 \pi} \sigma}e^{-(x-\mu)^2/(2\sigma^2)}\,, 
 \quad -\infty < x < \infty \,.
$$
We write $X \sim \mathsf{N}(\mu, \sigma^2)$. 
::: 

For $X\sim \mathsf{N}(\mu,\sigma^2),$ it can be shown that $\E(X) = \mu$ and $\Var(X) = \sigma^2,$ that is, $\mu$ is the *mean* and $\sigma^2$ is the *variance* of $X$. The pdf forms a bell-shaped curve that is symmetric about $\mu,$ as illustrated in @fig-normals-diff-mean. The value $\sigma$ (*standard deviation*) is the distance from $\mu$ to the inflection points of the curve. As $\sigma$ increases, the dispersion in the density increases, as illustrated in @fig-normals-diff-sd. Thus, the distribution's position (location) and spread depend on $\mu$ and $\sigma$.  

```{r}
#| label: fig-normals-diff-mean
#| fig-cap: "The pdfs of two normal rvs, $X_1 \\sim \\mathsf{N}(-2, 1)$ and $X_2 \\sim \\mathsf{N}(2, 1),$ with *different means* and the same standard deviations."
#| echo: false
#| warning: false
#| message: false
norm1.dist <- data.frame(x = seq(-8, 8, by =0.01)) |> 
 mutate(X1 = dnorm(x = x, mean = -2, sd = 1),
        X2 = dnorm(x = x, mean = 2, sd = 1)) |>
 gather(key = "X", value = "density", -x)
norm1.dist$X <- factor(norm1.dist$X, levels = c("X1", "X2"), labels = c("-2", "2"))
ggplot(norm1.dist, aes(x = x, y = density, color = X, linetype = X)) +
 geom_line(linewidth = lsz) +
 labs(y = TeX("$f(x;\\mu, \\sigma)$"), x = TeX("$x$"), color = TeX("$\\mu$"), linetype = TeX("$\\mu$")) + theme_ur
```  

```{r}
#| label: fig-normals-diff-sd
#| fig-cap: "The pdfs of two normal rvs, $X_1 \\sim \\mathsf{N}(0, 9)$ and $X_2 \\sim \\mathsf{N}(0, 1),$ with the same means and *different standard deviations*."
#| echo: false
#| warning: false
#| message: false
norm2.dist <- data.frame(x = seq(-8, 8, by =0.01)) |> 
 mutate(X1 = dnorm(x = x, mean = 0, sd = 3),
        X2 = dnorm(x = x, mean = 0, sd = 1)) |>
 gather(key = "X", value = "density", -x)
norm2.dist$X <- factor(norm2.dist$X, levels = c("X1", "X2"), labels = c("3", "1"))
ggplot(norm2.dist, aes(x = x, y = density, color = X, linetype = X)) +
 geom_line(linewidth = lsz) +
 labs(y = TeX("$f(x;\\mu, \\sigma)$"), x = TeX("$x$"), color = TeX("$\\sigma$"), linetype = TeX("$\\sigma$")) + theme_ur
```  

::: {#def-standard-normal}
## Standard normal distribution

We say that $X$ has a standard normal distribution if $\mu=0$ and $\sigma = 1$ and we will usually denote standard normal rvs by $$Z \sim \mathsf{N}(0,1)$$ (why $Z$? tradition!^["Traditions, traditions... Without our traditions, our lives would be as shaky as a fiddler on the roof!" [[https://www.youtube.com/watch?v=gRdfX7ut8gw](https://www.youtube.com/watch?v=gRdfX7ut8gw)].]). We denote the cdf of the standard normal by $$\Phi(z) = P(Z \leq z)$$ and write $\varphi = \Phi'$ for its density function. 
:::

::: {.callout-important}
## Useful facts about normal variates

1. If $X \sim \mathsf{N}(\mu, \sigma^2),$ then $$Z = (X - \mu) / \sigma  \sim \mathsf{N}(0,1).$$
2. If $Z \sim \mathsf{N}(0, 1),$ then $$X = \mu + \sigma Z \sim \mathsf{N}(\mu, \sigma^2).$$
3. If $X_i \sim \mathsf{N}(\mu_i, \sigma_i^2)$ for $i = 1, \dots, n$ are independent rvs, then 
$$\sum_{i=1}^{n} X_i \sim \mathsf{N} \left( \sum_{i=1}^{n} \mu_i, \sum_{i=1}^{n} \sigma_i^2 \right) \,.$$  
::: 

::: {.callout-warning}
## Variances add

In particular, for differences of independent rvs $X_1 \sim \mathsf{N}(\mu_1, \sigma_1^2)$ and $X_2 \sim \mathsf{N}(\mu_2, \sigma_2^2)$ then the variances add:
$$ X_1 - X_2 \sim \mathsf{N}(\mu_1 - \mu_2, \sigma_1^2 + \sigma_2^2) \,.$$  
:::

Probabilities $P(a \leq X \leq b)$ are found by converting the problem in $X \sim \mathsf{N}(\mu, \sigma^2)$ to the *standard normal* distribution $Z \sim \mathsf{N}(0, 1)$ whose probability values $\Phi(z) = P(Z\leq z)$ can then be looked up in a table. From (1.) above, 
$$
\begin{aligned}
   P(a < X < b) &= P\left( \frac{a-\mu}{\sigma} < Z < \frac{b-\mu}{\sigma} \right) \\ 
    &= \Phi \left( \frac{b-\mu}{\sigma}\right) - \Phi\left(\frac{a-\mu}{\sigma}\right) \,.
\end{aligned}
$$
This process is often referred to as *standardising* (the normal rv).  

::: {#exm-norm-rt}

Let $X \sim \mathsf{N}(5, 9)$ and find $P(X \geq 5.5)$. 

$$
\begin{aligned}
   P(X \geq 5.5) &= P\left(Z \geq \frac{5.5 - 5}{3}\right) \\
    &= P(Z \geq 0.1667) \\
    &= 1 - P(Z \leq 0.1667) \\
    &= 1 - \Phi(0.1667) \\
    &= 1 - 0.5662 \\
    &= 0.4338\,,
\end{aligned}
$$
where we look up the value of $\Phi(z) = P(Z\leq z)$ in a table of standard normal curve areas.

The probability corresponds to the shaded area under the normal density $\varphi(x) = \Phi'(x)$ corresponding to $x \geq 5.5$ (see @fig-example-norm-geq). To calculate this area, we can also use the `R` code: `pnorm(5.5, mean = 5, sd = 3, lower.tail = FALSE)`.  

```{r}
#| label: fig-example-norm-geq
#| fig-cap: "The normal density $\\mathsf{N}(5,9)$ with the (one-sided) interval shaded in blue that corresponds to the probability $P(X \\geq 5.5)$."
#| echo: false
#| warning: false
#| message: false
ggplot(NULL, aes(c(-4, 14))) + 
 geom_line(stat = "function", fun = dnorm, args = list(mean = 5, sd = 3), xlim = c(-4, 14), size = lsz) +
 geom_area(stat = "function", fun = dnorm, args = list(mean = 5, sd = 3), fill = "blue", xlim = c(5.5, 14), alpha = 0.3) + 
 geom_segment(aes(x = 5.5, y = 0, xend = 5.5, yend = 0.131), linetype = "dashed") + 
 theme(axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), axis.text.x = element_blank()) + 
 annotate("text", x = 5.5, y = -0.005, label = TeX("$x \\geq 5.5$"), size = tsz, color = "blue") + 
 annotate("text", x = 10, y = 0.1, label = TeX("$X \\sim N(5,9)$"), size = tsz)
```
:::


::: {#exm-norm-dt}

Let $X \sim \mathsf{N}(5, 9)$ and find $P(4 \leq X \leq 5.25)$. 

$$
\begin{aligned}
   P(4 \leq X \leq 5.25) &= P\left(\frac{4-5}{3} \leq Z \leq \frac{5.25-5}{3}\right) \\
   &= P(-0.3333 \leq Z \leq 0.0833) \\
   &= \Phi(0.0833) - \Phi(-0.3333) \\
   &= 0.5332 - 0.3694 \\
   &= 0.1638\,.
  \end{aligned}
$$
where we look up the value of $\Phi(z) = P(Z\leq z)$ in a table of standard normal curve areas. 

The probability corresponds to the shaded area under the normal density $\varphi(x) = \Phi'(x)$ corresponding to $4 \leq x \leq 5.25$ (see @fig-example-norm-interval). To calculate this area, we can use the `R` code: `pnorm(5.25, mean = 5, sd = 3) - pnorm(4, mean = 5, sd = 3)`. 

```{r}
#| label: fig-example-norm-interval
#| fig-cap: "The normal density $\\mathsf{N}(5,9)$ with the (two-sided) interval shaded in blue that corresponds to the probability $P(4 \\leq X \\leq 5.25)$."
#| echo: false
#| warning: false
#| message: false
ggplot(NULL, aes(c(-4, 14))) + 
 geom_line(stat = "function", fun = dnorm, args = list(mean = 5, sd = 3), xlim = c(-4, 14), size = lsz) +
 geom_area(stat = "function", fun = dnorm, args = list(mean = 5, sd = 3), fill = "blue", xlim = c(4, 5.25), alpha = 0.3) + 
 geom_segment(aes(x = c(4, 5.25), y = 0, xend = c(4, 5.25), yend = c(0.125, 0.133)), linetype = "dashed") + 
 theme(axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), axis.text.x = element_blank()) + 
 annotate("text", x = 4.625, y = -0.005, label = TeX("$4 \\leq x \\leq 5.25$"), size = tsz, color = "blue") + 
 annotate("text", x = 10, y = 0.1, label = TeX("$X \\sim \\textbf{N}(5,9)$"), size = tsz)
```
:::

::: {.callout-important}
## Empirical rule ($68-95-99.7$ rule)

For samples from a normal distribution, the percentage of values that lie within one, two, and three standard deviations of the mean are $68.27\%,$ $95.45\%,$ and $99.73\%,$ respectively. That is, for $X \sim \mathsf{N}(\mu, \sigma^2),$ 
$$
P(\mu - 1 \sigma \leq X \leq \mu + 1 \sigma ) \approx 0.6827\,,
$$
$$
P(\mu - 2 \sigma \leq X \leq \mu + 2 \sigma ) \approx 0.9545\,,
$$
$$
P(\mu - 3 \sigma \leq X \leq \mu + 3 \sigma ) \approx 0.9973\,.
$$
For a normal population, nearly all the values lie within "three sigmas" of the mean.
:::

## Student's $\mathsf{t}$ distribution {#sec-t-distribution}

Student's $\mathsf{t}$ distribution gets its peculiar name as it was first published under the pseudonym  "Student".^[William Sealy Gosset (1876--1937) wrote under the pseudonym "Student" [[https://mathshistory.st-andrews.ac.uk/Biographies/Gosset/](https://mathshistory.st-andrews.ac.uk/Biographies/Gosset/)].] This bit of obfuscation was to protect the identity of his employer,^[Gosset invented the t-test to handle small samples for quality control in brewing, specifically for the Guinness brewery in Dublin [[https://www.wikiwand.com/en/Guinness_Brewery](https://www.wikiwand.com/en/Guinness_Brewery)].] and thereby vital trade secrets, in a highly competitive and lucrative industry.  

::: {#def-t-dist}
## Student's $\mathsf{t}$ distribution

A continuous rv $X$ has a $\mathsf{t}$ distribution with parameter $\nu > 0,$ if $X$ has pdf
$$
f(x; \nu) = \frac{\Gamma\left(\tfrac{\nu+1}{2}\right)}{\sqrt{\nu \pi} \Gamma \left(\tfrac{\nu}{2}\right)} \left( 1 + \tfrac{x^2}{\nu} \right)^{- \frac{\nu+1}{2}} \,, \quad -\infty < x < \infty\,.
$$
We write $X \sim \mathsf{t}(\nu)$. Note $\Gamma$ is the standard gamma function.^[The gamma function is defined by $\Gamma(z) = \int_0^\infty x^{z-1}e^{-x} dx$ when the real part of $z$ is positive. For any positive integer $n,$ $\Gamma(n) = (n-1)!$ and for half-integers $\Gamma(\tfrac{1}{2} + n) = \frac{(2n)!}{4^n n!} \sqrt{\pi}$.]
:::

The density for $\mathsf{t}(\nu)$ for several values of $\nu$ are plotted below in @fig-example-t-dist. 

```{r}
#| label: fig-example-t-dist
#| fig-cap: "The density for $\\mathsf{t}(\\nu)$ for several values of $\\nu$ (df)."
#| echo: false
#| warning: false
#| message: false
t.dist <- data.frame(t = seq(-4, 4, 0.1)) |> 
 mutate(df1 = dt(x = t, df = 1), 
        df2 = dt(x = t, df = 2), 
        df5 = dt(x = t, df = 5), 
        df1000 = dt(x = t, df = 1000)) |> 
 gather(key = "df", value = "density", -t)
t.dist$df <- factor(t.dist$df, levels = c("df1", "df2", "df5", "df1000"), labels = c(1, 2, 5, 1000))
ggplot(t.dist, aes(x = t, y = density, color = df)) +
 geom_line(linewidth = lsz) +
 labs(y = TeX("$f(x;\\nu)$"), x = TeX("$x$"), color = TeX("$\\nu$ (df)")) + theme_ur
```

::: {.callout-important}
## Properties of $\mathsf{t}$ distributions {#facts-t}

1. The density for $\mathsf{t}(\nu)$ is a bell-shaped curve centred at $0$.
2. The density for $\mathsf{t}(\nu)$ is more spread out than the standard normal density (i.e., it has "fatter tails" than the normal).
3. As $\nu \to \infty,$ the spread of the corresponding $\mathsf{t}(\nu)$ density converges to the standard normal density (i.e., the spread of the $\mathsf{t}(\nu)$ density decreases relative to the standard normal).  

If $X \sim \mathsf{t}(\nu),$ then $\E[X] = 0$ for $\nu > 1$ (otherwise the mean is undefined).  
:::

::: {.callout-note}
## Cauchy distribution

A $\mathsf{t}$ distributions with $\nu = 1$ has pdf 
$$f(x) = \frac{1}{\pi (1 + x^2)}\,,$$
and we call this the Cauchy distribution. 
:::


## $\chi^2$ distribution {#sec-chisq-distribution}

The $\chi^2$ distribution arises as the distribution of a sum of the squares of $\nu$ independent standard normal rvs.  

::: {#def-chisq-dist}
## $\chi^2$ distribution

A continuous rv $X$ has a $\chi^2$ distribution with parameter $\nu \in \mathbf{N}_{>},$ if $X$ has pdf
\begin{equation*}
f(x; \nu) = \frac{1}{2^{\nu/2} \Gamma(\nu/2)} x^{(\nu/2)-1} e^{-x/2} \,, 
\end{equation*}
with support $x \in (0, \infty)$ if $\nu=1,$ otherwise $x \in [0, \infty)$. We write $X \sim \chi^2(\nu)$. 
:::  

The pdf $f(x; \nu)$ of the $\chi^2(\nu)$ distribution depends on a positive integer $\nu$ referred to as the df. The densities for several values of $\nu$ are plotted below in @fig-example-chisq-dist. The density $f(x;\nu)$ is positively skewed, i.e., the right tail is longer, so the mass is concentrated to the figure's left in @fig-example-chisq-dist. The distribution becomes more symmetric as $\nu$ increases. We denote critical values of the $\chi^2(\nu)$ distribution by $\chi^2_{\alpha, \nu}$.

```{r}
#| label: fig-example-chisq-dist
#| fig-cap: "The density for $\\chi^2(\\nu)$ for several values of $\\nu$ (df)."
#| echo: false
#| warning: false
#| message: false
chisq.dist <- data.frame(chisq = 0:7000 / 100) |> 
 mutate(df1 = dchisq(x = chisq, df = 1), 
        df2 = dchisq(x = chisq, df = 2),
        df3 = dchisq(x = chisq, df = 3),
        df4 = dchisq(x = chisq, df = 4),
        df6 = dchisq(x = chisq, df = 6),
        df9 = dchisq(x = chisq, df = 9)) |>
 gather(key = "df", value = "density", -chisq)
chisq.dist$df <- factor(chisq.dist$df, levels = c("df1", "df2", "df3", "df4", "df6", "df9"), labels = c(1, 2, 3, 4, 6, 9))
ggplot(chisq.dist, aes(x = chisq, y = density, color = df)) +
 geom_line(linewidth = lsz) + ylim(c(0,.6)) + xlim(c(0,15)) +
 labs(y = TeX("$f(x;\\nu)$"), x = TeX("$x$"), color = TeX("$\\nu$ (df)")) + theme_ur
```

:::{.callout-warning} 
## Skew

Unlike the normal and $t$ distributions, the $\chi^2$ distribution is not symmetric! This means that critical values, e.g., $$\chi^2_{.99, \nu} \quad \text{and}\quad \chi^2_{0.01,\nu}\,,$$ are **not** equal. Hence, it will be necessary to look up both values for CIs based on $\chi^2$ critical values.  
:::

If $X \sim \chi^2(\nu),$ then $\E[X] = \nu$ and $\Var[X] = 2\nu$.  

## $\mathsf{F}$ distribution {#sec-F-distribution}

The $\mathsf{F}$ distribution ("F" for Fisher) arises as a test statistic when comparing population variances and in the analysis of variance (see @sec-anova). 

::: {#def-F-dist}
## $\mathsf{F}$ distribution
A continuous rv $X$ has an $\mathsf{F}$ distribution with df parameters $\nu_1$ and $\nu_2,$ if $X$ has pdf
$$
 f(x; \nu_1, \nu_2) = 
    \frac{\Gamma\left(\frac{\nu_1+\nu_2}{2}\right) \nu_1^{\nu_1/2} \nu_2^{\nu_2/2}}
 {\Gamma\left(\frac{\nu_1}{2}\right) \Gamma\left(\frac{\nu_2}{2}\right)} 
 \frac{x^{\nu_1/2 - 1}}{(\nu_2+\nu_1 x)^{(\nu_1+\nu_2)/2}} \,.
$$
:::

The pdf $f(x; \nu_1, \nu_2)$ of the $\mathsf{F}(\nu_1, \nu_2)$ distribution depends on two positive integers $\nu_1$ and $\nu_2$ referred to, respectively, as the numerator and denominator df. The density is plotted below for several combinations of $(\nu_1, \nu_2)$ in @fig-example-F-dist.

```{r}
#| label: fig-example-F-dist
#| fig-cap: "The density for $\\mathsf{F}(\\nu_1, \\nu_2)$ for several combinations of $(\\nu_1, \\nu_2)$."
#| echo: false
#| warning: false
#| message: false
f.dist <- data.frame(f = 0:1000 / 100) |> 
 mutate(df_1_1 = df(x = f, df1 = 1, df2 = 1),
        df_2_1 = df(x = f, df1 = 2, df2 = 1),
        df_5_2 = df(x = f, df1 = 5, df2 = 2),
        df_10_2 = df(x = f, df1 = 10, df2 = 2),
        df_100_100 = df(x = f, df1 = 100, df2 = 100)) |>
 gather(key = "df", value = "density", -f)
f.dist$df <- factor(f.dist$df, levels = c("df_1_1", "df_2_1", "df_5_2", "df_10_2", "df_100_100"), labels = c("(1, 1)", "(2, 1)", "(5, 2)", "(10, 2)", "(100, 100)"))
ggplot(f.dist, aes(x = f, y = density, color = df)) +
 geom_line(size = lsz) + ylim(c(0,2.1)) + xlim(c(0,3)) +
 labs(y = TeX("$f(x;\\nu_1, \\nu_2)$"), x = TeX("$x$"), color = TeX("$(\\nu_1, \\nu_2)$")) + theme_ur
```

:::{.callout-tip}
## Where do the terms numerator and denominator df come from?

The $\mathsf{F}$ distribution is related to ratios of $\chi^2$ rvs, as captured in @thm-F-dist-chisq.
:::



::: {#thm-F-dist-chisq}
## Ratio of $\chi^2$ rvs

If $X_1 \sim \chi^2(\nu_1)$ and $X_2 \sim \chi^2(\nu_2)$ are independent rvs, then the rv 
$$
 F = \frac{X_1 / \nu_1}{X_2 / \nu_2} \quad \sim \mathsf{F}(\nu_1,\nu_2)\,,
$$
that comprises the ratio of two $\chi^2$ rvs divided by their respective df has an $\mathsf{F}(\nu_1, \nu_2)$ distribution. 
:::
 
