```{r setup, include=FALSE}
#  bookdown::render_book("index.Rmd", "bookdown::gitbook", output_dir = "docs/")

# packages
library(bookdown)
library(RColorBrewer)
library(readr)
library(plotly)
library(tidyverse)
library(reshape2)
library(latex2exp)
library(svglite)
library(knitr)
library(kableExtra)
library(formattable)
library(DT)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE, comment = "", dev = "svglite", fig.width=8) #, dev = "svglite", fig.width=8
#plotly::config(plot_ly(), displaylogo = FALSE, mathjax = "cdn")
```

# Course Guide {-#cg}

## Organisation {-#cg-organization}

The MA22004 module runs for 11 teaching weeks and is worth 20 SCQF credits (equal to 10 ECTS points). All organisation and teaching will be carried out by:

--------------------
Dr Eric Hall
[ehall001@dundee.ac.uk](mailto:ehall001@dundee.ac.uk)

Mathematics Division
Room TBA,
Fulton Building

01382 TBA
--------------------

This course uses [Blackboard Ultra](my.dundee.ac.uk) (look for course `MA22004_SEM0000_2021`) for communicating all announcements/deadlines and also for running online meetings.

You should make an appointment to see Dr Hall if you have a problem regarding the course. You may also bring matters of concern about the course  to the attention of the Mathematics Division Staff/Student Committee, which meets once each semester. A volunteer from Level 2 Mathematics will act as class representative to sit on the Staff/Student Committee; their name will be posted on Ultra.

## Timetable {-#cg-timetable}

> ⚠️ Due to COVID19, these plans may be subject to change. 

The delivery of this module consists of a blend of synchronous and asynchronous content delivered both in-person and online. On an average week there will be seven planned teaching and learning activities.

```{r table-activities, echo=FALSE, warning = FALSE, message = FALSE, fig.cap = "Learning activities."}
kable(tibble(read_csv("data/activities.csv")))
```

The anticipated student effort is 200 hours over the length of the module. You are expected to be "present" for all synchronous timetabled activities with the exception of the online office hours which are optional. You may engage with the asynchronous material at your own pace, keeping in mind to meet any deadlines for engagement and/or attainment that will be posted to Ultra and discussed. 

## Pre-requisites {-#cg-prerecs}

In order to take this course you must have passed module MA12003 or an equivalent qualification.

## Syllabus {-#cg-syllabus}

Sampling Distributions 
: Mean and standard deviation of samples, sampling from a single population, sampling from two populations, shape of sampling distributions. Normal distribution, $\chi^2$-square distribution, F-distribution.

Hypothesis tests 
: Null and Alternate hypotheses, inferences, confidence intervals, estimating means, proportions and standard deviations.

Linear Regression
: Least squares, assessing usefulness of a model, using a model. 

Industrial Quality Control
: Control Charts, acceptance sampling.

R software package 
: Appropriate use of computational software to carry out statistical and probabilistic calculations.

## Recommended Books {-#cg-books}

In addition to the course notes, here are some text books you may wish to consult.

> ⚠️ You do not need to purchase these books.

+ Devore, *Probability and Statistics for Engineering and the Sciences,* Cengage learning, 2011. [§6-10, 12]
+ DeGroot and Schervish, *Probability and Statistics,* Addison-Wesley, 2001. [§7-10]
+ Rice, *Mathematical statistics and data analysis,* Cengage Learning, 2006. [§6-12]
+ Wasserman, *All of Statistics,* Springer-Verlag, 2004. [Concise general reference]

## Assessment {-#cg-assessment}

The module will be *continuously* assessed using coursework and examinations. Deadlines as well as test dates will be posted on Ultra and announcements made in the class hours. The module assessment weighting is as follows. 

```{r table, echo=FALSE}
df <- tibble(Assessment=c("Assignments", "Midterm Exam 1", "Midterm Exam 2", "Final Exam"),
             Weight=c("20%", "20%", "20%", "40%"))
kable(df)
```

### Coursework {-#cg-assessment-coursework}

The coursework will be assessed primarily through:

- six hand-in laboratory reports and 
- weekly engagement with the reading material using [Perusall](https://perusall.com/). 

There will also be alternative means of demonstrating your mastery of course material through:

- one (group) lab presentation and
- short seminar quizzes (announced in advance).

### Examinations {-#cg-assessment-exams}

The **Midterm Exams** will be computer-assessed and will be one (1) hour in scope. These will likely be held in week 4 and week 8. 

The **Final Exam** will be a two (2) hour hand-written exam. You will be given one additional hour to scan and upload your exam script using [Gradescope](https://www.gradescope.com/). This process will be thoroughly discussed and practiced with a dummy exam in advance of a real submission. The Final Exam will be given in the last week of term (i.e., during week 11).

To pass this module it is necessary to obtain an overall grade of at least `D3` in the overall assessment **and** to obtain a grade of at least `M1` for the exam **and** to obtain a grade of at least `M1` for the coursework.

> ⚠️ For those who fail the module, there may be an opportunity to take a two-hour resit examination paper at the next available exam diet. However, unless you have mitigating circumstances, if you fail to achieve a module grade of `CF` or above at first attempt then you may not be permitted to resit the exam. Also, unless you have mitigating circumstances, any pass after a resit will be capped at a grade of `D3` regardless of the weighted average mark obtained. Resit marks are based on the resit exam only. 

## Your Commitment {-#cg-commitment}

You should attend all synchronous timetabled sessions except on medical grounds or with the special permission of Dr Hall. If you are unable to attend the degree examination or complete elements of the coursework on time then you should inform Dr Hall and submit a medical certificate. Medical certificates should be submitted to your School Office as soon as possible after the abscence.

> ⚠️ You must also  submit a Mitigating Circumstances form to explain which aspects of assessment have been affected by your abscence.

A Medical Certificate will not be taken into account unless a Mitigating Circumstances form that refers to the medical certificate has also been completed.

## Approved Calculators {-#cg-calcs}

The only types of calculators that have been approved for use in assessments in the School of Engineering, Physics and Mathematics are the Casio FX83 and the Casio FX85.

## Study Support {-#cg-support}

If you are  having difficulty with the course you are encouraged to seek help at an early stage by making an appointment to meet with Dr Hall. You may also obtain additional help from the Maths Base (see Ultra for details).

## Disability {-#cg-disability}

The University of Dundee is committed to making reasonable, effective and appropriate accommodations to meet the needs of students with disabilities and help create an inclusive and barrier-free campus. If you are in need of accommodation for a documented disability, then you are advised to register with [Disability Services](https://www.dundee.ac.uk/disabilityservices/). Please communicate any needs you may have directly with Dr Hall and ahead of time to ensure that any accommodations might be managed in a timely manner. 

## Academic Honesty {-#cg-honesty}

Since the integrity of the academic enterprise of any higher education institution requires honesty in scholarship and research, academic honesty is required of all students at the University of Dundee. Academic dishonesty includes but is not limited to: cheating, fabrication, plagiarism, and facilitating dishonesty. Appropriate sanctions may be imposed on any student who has committed an act of academic dishonesty and ignorance of such standards is not normally sufficient evidence of lack of intent. For more information about what constitutes academic dishonesty, please see the [*Code of Practice on Academic Misconduct by Students*](https://www.dundee.ac.uk/qf/documents/details/academic-misconduct.php).

## End of Module Questionaire {-#cg-questionaire}

At the end of each section of the module you will be asked to complete a confidential questionnaire regarding the content and presentation of the module. This is an important element in the University's Academic Standards procedures. 