# Module Guide {-#cg}

<!-- ::: {.importantblock data-latex=""}  -->
<!-- Due to COVID19, all plans may be subject to change.  -->
<!-- :::  -->

## Organisation {-#cg-organization}

All organisation and teaching will be carried out by:

:::{.noteblock data-latex=""}
Dr Eric Hall  

[ehall001@dundee.ac.uk](mailto:ehall001@dundee.ac.uk)

+44 1382 384632

Mathematics Division

School of Science and Engineering

Fulton G10

:::

A tutor, Mr Miloš Mičík, will support the module. Miloš will assist with the workshops and can answer questions about the labs/`R`/RStudio posted to MS Teams. 

This module uses MyDundee (a.k.a Blackboard Ultra)^[Visit [https://my.dundee.ac.uk](https://my.dundee.ac.uk) and search for `MA22004_SEM0100_2223`.] for communicating all announcements/deadlines. This module will use Gradescope^[With Blackboard integration, for more information about Gradescope, see [https://www.gradescope.com/](https://www.gradescope.com/).] for submission of lab reports and to deliver feedback on class tests and Perusall^[Available through MyDundee.] for collaborative peer learning focusing on the weekly reading materials. An **MS Teams channel will be available for contacting the module tutor** and for asynchronous communication. 

If you have a problem regarding the module, then you should make an appointment to speak with Dr Hall. You may also bring matters of concern about the module to the attention of the Mathematics Division Student-Staff Liaison Committee (SSLC). A volunteer from Level 2 Mathematics will act as a representative to sit on the SSLC (see Ultra for contact details). 

## Timetable {-#cg-timetable}

::: {.warningblock data-latex=""} 
Please remember to check in to synchronous activities using **SEAtS**, the University's attendance monitoring system. Further information is available at:

[https://www.dundee.ac.uk/guides/check-your-classes-seats-mobile-app](https://www.dundee.ac.uk/guides/check-your-classes-seats-mobile-app). 
::: 

This module runs for 11 teaching weeks and is worth 20 SCQF credits (equivalently, 10 ECTS points). The delivery of this module consists of a blend of synchronous and asynchronous content delivered in-person and online. On a *typical* week, there will be five planned teaching and learning activities.

```{r table-activities, echo=FALSE, warning = FALSE, message = FALSE, fig.cap = "Learning activities."}
kbl(tibble(read_csv("data/activities.csv")), booktabs = T) %>%
 kable_styling(latex_options = c("striped", "hold_position"), )
```

Some weeks will include class tests or other timetabled activities in place of the computer labs. The anticipated student effort is 200 hours over the length of the module. **You are expected to be present for all synchronous timetabled activities and check-in using SEAtS.** You may engage with the asynchronous material at your own pace but must meet deadlines for engagement and attainment that will be posted to Ultra and discussed. 

## Pre-requisites {-#cg-prerecs}

You must have passed module MA12003 Statistics I or equivalent to take this module.

## Syllabus {-#cg-syllabus}

Sampling Distributions 
: Mean and standard deviation of samples, sampling from a single population, sampling from two populations, shape of sampling distributions. Normal distribution, $\chi^2$-square distribution, F-distribution.

Hypothesis tests 
: Null and Alternate hypotheses, inferences, confidence intervals, estimating means, proportions and standard deviations.

Linear Regression
: Least squares, assessing usefulness of a model, using a model.

Industrial Quality Control
: Control Charts, acceptance sampling.

R software package 
: Appropriate use of computational software to carry out statistical and probabilistic calculations.

## Recommended Books {-#cg-books}

In addition to the lecture notes, here are some textbooks you may wish to consult.

+ *Probability and Statistics for Engineering and the Sciences,* [@Devore:2016st, §6-10, 12, 14, 16]
+ *Probability and Statistics,* [@DeGrootSchervish:2001ps, §7-10]
+ *Introductory Statistics with R,* [@Dalgaard:2008st, §6-7]
+ *Mathematical statistics and data analysis,* [@Rice:2006da, §6-12]
+ *All of Statistics,* [@Wasserman:2013as] a concise general reference
+ *R for data science,* [@WickhamGrolemund:2017or]

I would also recommend Prof David Spiegelhalter's new book *The Art of Statistics: Learning from Data* [@Spiegelhalter:2020ps].

::: {.noteblock data-latex=""}
You do not need to purchase any books for this course. Check the *Library Resource List* on MyDundee. 
:::  

## Assessment {-#cg-assessment}

The module will be *continuously* assessed using coursework and class tests. Deadlines and test dates will be posted on MyDundee, and announcements will be made in the seminars. The module assessment weighting is as follows. 

```{r table, echo=FALSE}
df <- tibble(Assessment=c("Labs", "Peer Learning", "Class Test 1", "Class Test 2"),
             Weight=c("35% (7 @ 5% each)", "5%", "30%", "30%"))
kbl(df, booktabs = T, align = c("l", "r")) %>%
 kable_styling(latex_options = c("striped", "hold_position"))
```

### Coursework {-#cg-assessment-coursework}

Assessed coursework includes:

- seven hand-in laboratory reports and 
- weekly peer learning engagement using Perusall. 

Please participate by providing positive and constructive feedback. 

All lab reports will be submitted using Gradescope.

### Class Tests {-#cg-assessment-exams}

**Class Test 1** and **Class Test 2** will be 60 minutes long. A peer learning activity will follow each class test. 

All class tests will be hand-written. Feedback will be delivered using Gradescope. You are encouraged to use the Gradescope facilities to read and ask questions about the feedback if you have any. 

For those who fail the module, there may be an opportunity to take a two-hour resit examination (typically an oral examination) at the next available exam diet. 

:::{.warningblock data-latex=""}
Resit marks are based on the resit exam only.
:::

Unless you have mitigating circumstances, if you fail to achieve a module grade of `CF` or above at first attempt, you may not be permitted to resit the exam. Also, unless you have mitigating circumstances, any pass after a resit will be capped at a grade of `D3` regardless of the weighted average mark.  

## Your Commitment {-#cg-commitment}

You should attend all synchronous timetabled sessions except on medical grounds or with the special permission of Dr Hall. If you cannot participate in the class tests or complete elements of the coursework on time, then you should inform Dr Hall and submit a medical certificate. Medical certificates should be submitted to your School Office as soon as possible after the absence. You must also submit a Mitigating Circumstances form to explain which aspects of the assessment have been affected by your absence. A Medical Certificate will only be considered if accompanied by a completed Mitigating Circumstances form that refers to the medical certificate.

## Approved Calculators {-#cg-calcs}

The Casio FX83 and the Casio FX85 are the only calculators approved for use in assessments in the School of Engineering, Physics and Mathematics.

## Study Support {-#cg-support}

If you are having difficulty with the module, you are encouraged to seek help at an early stage by making an appointment with Dr Hall. You may also obtain additional support from the Maths Cafe (there ).

## Disability {-#cg-disability}

The University of Dundee is committed to making reasonable, effective and appropriate accommodations to meet the needs of students with disabilities and to create an inclusive and barrier-free campus. If you require accommodation for a documented disability, then you are advised to register with Disability Services.^[Website available at [https://www.dundee.ac.uk/disabilityservices/](https://www.dundee.ac.uk/disabilityservices/).] Please communicate any needs you may have directly with Dr Hall as soon as possible to ensure timely management of any accommodations. 

## Academic Honesty {-#cg-honesty}

Honesty in scholarship and research is integral to the integrity of the academic enterprise of any higher education institution. Therefore, all students at the University of Dundee must understand and practice academic integrity. Academic dishonesty includes cheating, fabrication, plagiarism, and facilitating dishonesty. Cases of academic dishonesty will be subject to appropriate sanctions, and ignorance of such standards is insufficient evidence of lack of intent. Please see the *Code of Practice on Academic Misconduct by Students*^[Available at [https://www.dundee.ac.uk/qf/documents/details/academic-misconduct.php](https://www.dundee.ac.uk/qf/documents/details/academic-misconduct.php).] for more information about what constitutes academic dishonesty.

## Module Questionaires {-#cg-questionaire}

You will have the opportunity to periodically complete a confidential questionnaire regarding the content and presentation of the module. These questionnaires form an essential element in the University's Academic Standards procedures. Thank you in advance for your cooperation. 
