# Course Guide {-#cg}

## Organisation {-#cg-organization}

All organisation and teaching will be carried out by:

:::{.noteblock data-latex=""}
Dr Eric Hall   

[ehall001@dundee.ac.uk](mailto:ehall001@dundee.ac.uk)

Mathematics Division   

Fulton G10

+44 (0)1382 384488
:::

The module will be supported by two tutors: Ms Lucy Brown and Mr Miloš Mičík. Lucy will assist with online workshops. Miloš will assist with the labs by answering questions about `R` and RStudio. 

This course uses MyDundee (a.k.a Blackboard Ultra)^[Visit [https://my.dundee.ac.uk](https://my.dundee.ac.uk) and search for course `MA22004_SEM0100_2122`.] for communicating all announcements/deadlines and also for running online meetings (we will use the Collaborate course room for online seminars and workshops). This course will use Gradescope^[Available at [https://www.gradescope.com/](https://www.gradescope.com/); "course entry code" at **Welcome > Overview of online platforms used in this course > Gradescope** on Blackboard Ultra.] for submission of continuous assessments and Perusall^[Available through Blackboard Ultra.] for collaborative engagement with reading materials. An **MS Teams channel will be available for contacting the module tutors** and for asynchronous communication. 

If you have a problem regarding the course, then you should make an appointment to speak with Dr Hall. You may also bring matters of concern about the course to the attention of the Mathematics Division Student-Staff Liaison Committee (SSLC). A volunteer from Level 2 Mathematics will act as class representative to sit on the SSLC (see Ultra for contact details). 

## Timetable {-#cg-timetable}

::: {.importantblock data-latex=""} 
Due to COVID19, these plans may be subject to change. 
::: 

This module runs for 11 teaching weeks and is worth 20 SCQF credits (equivalently, 10 ECTS points). The delivery of this module consists of a blend of synchronous and asynchronous content delivered both in-person and online. On an average week, there will be six planned teaching and learning activities.

```{r table-activities, echo=FALSE, warning = FALSE, message = FALSE, fig.cap = "Learning activities."}
kbl(tibble(read_csv("data/activities.csv")), booktabs = T) %>%
 kable_styling(latex_options = c("striped", "hold_position"), )
```

The anticipated student effort is 200 hours over the length of the module. You are expected to be "present" for all synchronous timetabled activities and to check-in using the University's attendance monitoring application. You may engage with the asynchronous material at your own pace, but please meet any deadlines for engagement and/or attainment that will be posted to Ultra and discussed. 

## Pre-requisites {-#cg-prerecs}

To take this course, you must have passed module MA12003 or equivalent.

## Syllabus {-#cg-syllabus}

Sampling Distributions 
: Mean and standard deviation of samples, sampling from a single population, sampling from two populations, shape of sampling distributions. Normal distribution, $\chi^2$-square distribution, F-distribution.

Hypothesis tests 
: Null and Alternate hypotheses, inferences, confidence intervals, estimating means, proportions and standard deviations.

Linear Regression
: Least squares, assessing usefulness of a model, using a model.

Industrial Quality Control
: Control Charts, acceptance sampling.

R software package 
: Appropriate use of computational software to carry out statistical and probabilistic calculations.

## Recommended Books {-#cg-books}

In addition to the course notes, here are some textbooks you may wish to consult.

+ *Probability and Statistics for Engineering and the Sciences,* [@Devore:2016st, §6-10, 12, 14, 16]
+ *Probability and Statistics,* [@DeGrootSchervish:2001ps, §7-10]
+ *Introductory Statistics with R,* [@Dalgaard:2008st, §6-7]
+ *Mathematical statistics and data analysis,* [@Rice:2006da, §6-12]
+ *All of Statistics,* [@Wasserman:2013as] a concise general reference

::: {.warningblock data-latex=""}
You do not need to purchase these books. Check the *Library Resource List* available on MyDundee. 
:::   

## Assessment {-#cg-assessment}

The module will be *continuously* assessed using coursework and class tests. Deadlines, as well as test dates, will be posted on Ultra and announcements made in the class hours. The module assessment weighting is as follows. 

```{r table, echo=FALSE}
df <- tibble(Assessment=c("Assignments", "Class Test 1", "Class Test 2", "Class Test 3"),
             Weight=c("20%", "20%", "20%", "40%"))
kbl(df, booktabs = T, align = c("l", "r")) %>%
 kable_styling(latex_options = c("striped", "hold_position"))
```

### Coursework {-#cg-assessment-coursework}

Assessed coursework includes:

- seven hand-in laboratory reports (the lowest non-zero lab report grade will be dropped)^[Yes---you must attempt and submit a report for each lab, but, on the bright side, there is some flexibility to have a bad day.] and 
- weekly engagement with the reading material using Perusall. 

The lab reports will be submitted using Gradescope.

<!-- There will also be alternative means of demonstrating your mastery of course material through: -->
<!-- - one (group) lab presentation and -->
<!-- - short seminar quizzes (announced in advance). -->

### Class Tests {-#cg-assessment-exams}

**Class Test 1** and **Class Test 2** will be 90 minutes in scope (with 30 minutes upload time). These will be in weeks 4 and 8. 

**Class Test 3** will be a 150 minutes in scope (with 30 minutes upload time). This will be in week 11 (i.e., during the last week of the term).

All class tests will be hand-written exams that you submit using Gradescope. This process will be thoroughly discussed and trialed with a dummy exam in advance of the real submission. 

To pass this module, you must:

- obtain an overall grade of at least `D3` in the overall assessment **and** 
- obtain a grade of at least `M1` for the class tests **and** 
- obtain a grade of at least `M1` for the coursework.

For those who fail the module, there may be an opportunity to take a two-hour resit examination (typically an oral examination) at the next available exam diet. 

:::{.warningblock data-latex=""}
Resit marks are based on the resit exam only.
:::

Unless you have mitigating circumstances, if you fail to achieve a module grade of `CF` or above at first attempt, then you may not be permitted to resit the exam. Also, unless you have mitigating circumstances, any pass after a resit will be capped at a grade of `D3` regardless of the weighted average mark obtained.  

## Your Commitment {-#cg-commitment}

You should attend all synchronous timetabled sessions except on medical grounds or with the special permission of Dr Hall. If you are unable to attend the degree examination or complete elements of the coursework on time, then you should inform Dr Hall and submit a medical certificate. Medical certificates should be submitted to your School Office as soon as possible after the absence. You must also submit a Mitigating Circumstances form to explain which aspects of assessment have been affected by your absence. A Medical Certificate will only be taken into account if accompanied by a completed Mitigating Circumstances form that refers to the medical certificate.

## Approved Calculators {-#cg-calcs}

The Casio FX83 and the Casio FX85 are the only calculators approved for use in assessments in the School of Engineering, Physics and Mathematics.

## Study Support {-#cg-support}

If you are having difficulty with the course, you are encouraged to seek help at an early stage by making an appointment with Dr Hall. You may also obtain additional help from the Maths Base (see Ultra for details).

## Disability {-#cg-disability}

The University of Dundee is committed to making reasonable, effective and appropriate accommodations to meet the needs of students with disabilities and to create an inclusive and barrier-free campus. If you require accommodation for a documented disability, then you are advised to register with Disability Services.^[Website available at [https://www.dundee.ac.uk/disabilityservices/](https://www.dundee.ac.uk/disabilityservices/).] Please communicate any needs you may have directly with Dr Hall and as soon as possible to ensure timely management of any accommodations. 

## Academic Honesty {-#cg-honesty}

Honesty in scholarship and research is integral to the integrity of the academic enterprise of any higher education institution. Therefore, all students at the University of Dundee must practice academic honesty. Academic dishonesty includes cheating, fabrication, plagiarism, and facilitating dishonesty. Cases of academic dishonesty will be subject to appropriate sanctions and ignorance of such standards is not sufficient evidence of lack of intent. Please see the *Code of Practice on Academic Misconduct by Students*^[Available at [https://www.dundee.ac.uk/qf/documents/details/academic-misconduct.php](https://www.dundee.ac.uk/qf/documents/details/academic-misconduct.php).] for more information about what constitutes academic dishonesty.

## End of Module Questionaire {-#cg-questionaire}

You will have the opportunity to complete a confidential questionnaire regarding the content and presentation of the module periodically. These questionnaires form an important element in the University's Academic Standards procedures. Thank you in advance for your cooperation. 
