---
title: "MA22004 – Statistics II"
author: "Dr Eric Hall"
date: 'Last updated: `r Sys.Date()`'
fontsize: 11pt
classoption: oneside
documentclass: book
site: bookdown::bookdown_site
github-repo: dundeemath/MA22004
url: https\://dundeemath.github.io/MA22004
bibliography: ma22004.bib
csl: assets/the-american-statistician.csl #www.zotero.org/styles
link-citations: true
link-bibliography: true
description: Course guide and course notes for MA22004 - Statistics and Probability
  II. Division of Mathematics, University of Dundee.
always_allow_html: yes
urlcolor: #4365e2
#include-before below needs to be commented out to compile pdf latex
include-before:
 - $\DeclareMathOperator{\E}{\mathbf{E}}$
 - $\DeclareMathOperator{\Var}{Var}$
 - $\DeclareMathOperator{\Cov}{Cov}$
 - $\DeclareMathOperator{\corr}{corr}$
 - $\newcommand{\se}{\mathsf{se}}$
 - $\DeclareMathOperator{\sd}{sd}$
---

```{r setup, include=FALSE}
#  bookdown::render_book("index.Rmd", "bookdown::gitbook", output_dir = "docs/")

# packages
library(bookdown)
library(RColorBrewer)
library(viridis)
library(scales)
library(readr)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(reshape2)
library(latex2exp)
library(svglite)
library(knitr)
library(kableExtra)
library(formattable)
library(DT)
library(dplyr)
library(datasets)
library(infer)
library(fontawesome)

options(knitr.kable.NA = '')

# ggplot theme
if (knitr::is_html_output()){
  theme_set(theme_gray(base_size = 12))
  knitr::opts_chunk$set(echo = TRUE, comment = "", fig.width = 8, fig.asp = 0.6, dev = "svglite", fig.align = "center")
} else {
  theme_set(theme_bw(base_size = 10))
  knitr::opts_chunk$set(echo = TRUE, comment = "", fig.width = 6, fig.asp = 0.6)
}
theme_ur <- theme(legend.justification = c(1,1), legend.position = c(1,1), legend.box.margin = margin(c(4, 4, 4, 4), unit = "pt"))
theme_lr <- theme(legend.justification = c(1,0), legend.position = c(1,0), legend.box.margin = margin(c(4, 4, 4, 4), unit = "pt"))
options(width = 90, knitr.kable.NA = '')
# opts <- options()  # save old options
# options(ggplot2.continuous.colour="viridis")
# options(ggplot2.continuous.fill = "viridis")
# scale_colour_discrete <- scale_colour_viridis_d
# data sets

lsz <- 1.0
tsz <- 4
data(iris)
data(trees)


#plotly::config(plot_ly(), displaylogo = FALSE, mathjax = "cdn")
#htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

# (PART\*) Module Documents {-}

# Welcome {-}

Welcome to MA22004 at the University of Dundee. 

This module covers the basics of statistical inference. The first part of this document contains information of a practical nature regarding the mechanics of the module. The second part of this document contains the content. The appendix contains a list of curated content for your to investigate. 

These are some trying times. 

Please try to stay healthy.

```{r about-me-html-only, echo = FALSE, child='about-me.Rmd', eval = knitr::is_html_output()}
```

***

These notes are available at [dundeemath.github.io/MA22004/](https://dundeemath.github.io/MA22004/) and also as a PDF (visit the page and click on the PDF icon to download). 

