---
title: "MA22004 - Statistics and Probability II"
author: "Dr Eric Hall"
date: "Last updated: `r Sys.Date()`"
fontsize: 11pt
classoption: oneside
documentclass: book
site: bookdown::bookdown_site
link-citations: yes
github-repo: dundeemath/MA22004
url: 'https\://dundeemath.github.io/MA22004'
bibliography: ma22004.bib
description: "Course guide and course notes for MA22004 - Statistics and Probability II. Division of Mathematics, University of Dundee."
always_allow_html: true
include-before:
  - '$\DeclareMathOperator{\E}{\mathbf{E}}$'
  - '$\DeclareMathOperator{\Var}{Var}$'
  - '$\DeclareMathOperator{\Cov}{Cov}$'
  - '$\DeclareMathOperator{\corr}{corr}$'
  - '$\newcommand{\se}{\mathsf{se}}$'
  - '$\DeclareMathOperator{\sd}{sd}$'
---

# (PART\*) Course Documents {-}
# Welcome {-}

Placeholder



<!--chapter:end:index.Rmd-->


# Course Guide {-#cg}

Placeholder


## Organisation {-#cg-organization}
## Timetable {-#cg-timetable}
## Pre-requisites {-#cg-prerecs}
## Syllabus {-#cg-syllabus}
## Recommended Books {-#cg-books}
## Assessment {-#cg-assessment}
### Coursework {-#cg-assessment-coursework}
### Examinations {-#cg-assessment-exams}
## Your Commitment {-#cg-commitment}
## Approved Calculators {-#cg-calcs}
## Study Support {-#cg-support}
## Disability {-#cg-disability}
## Academic Honesty {-#cg-honesty}
## End of Module Questionaire {-#cg-questionaire}

<!--chapter:end:guide.Rmd-->

# Lab Guide {-#labs}

You will learn about the statistical programming language `R` and the software RStudio by working through 7 interactive lab tutorials and by completing lab reports. 

These tutorials and all associated materials (templates, data sets, etc) are available as an `R` package at:

[https://github.com/dundeemath/MA22004labs](https://github.com/dundeemath/MA22004labs).

<!--chapter:end:labs.Rmd-->

# Deadlines {-#deadlines}

Below we have collected all the deadlines and important dates for the module. The three class tests (T1, T2, and T2) are on Wednesdays (W) from 13:00-15:00 and 14:00-17:00, respectively; we will practice (P) at uploading a test on Gradescope on Friday of week 1. Please keep in mind that you must engage with the weekly readings on Perusall **before** the seminar meeting on Wednesday afternoon. The lab reports (L1, L2, L3, L4, L5, L6, L7) are due on Thursday evenings at 17:00. 

Further details about the practice upload and the class tests will be circulated nearer in time to the relevant deadline.

```{r table-deadlines, echo=FALSE, warning = FALSE, message = FALSE, fig.cap = "An overview of all deadlines and important dates associated with the course."}
df <- tibble(read_csv("data/deadlines.csv")) %>% relocate(Reading, .after = Week)
kbl(df, booktabs = T) %>%
 kable_styling(latex_options = c("striped", "hold_position"))
```


<!--chapter:end:guide_deadlines.Rmd-->

# (PART\*) Course Notes {-}

# Preliminaries {-}

## Notation {-}

Uppercase roman letters, e.g., $X$, will typically denote random variables (rvs); lower case letters, e.g., $x$, will denote a particular value (observation) of a rv. Rvs have probability distributions. Distributions are typically characterized by *parameters* which are fixed real numbers. Parameters describe population characteristics that are often unknown and must be estimated from data. Statistical inference is a tool that will help us to do this.

:::{.warningblock}
Statistical models comprise both rvs and parameters. Be careful not to confuse them! 
:::

## Abbreviations {-}

```{r notation-abbrev, echo=FALSE, warning = FALSE, message = FALSE}
# datatable(tibble(read_csv("data/notation-abbrev.csv")), rownames = FALSE, #filter = 'bottom',
#           options = list(pageLength = 5, scrollX = T), autoHideNavigation = TRUE,
#           caption = "Collection of abbreviations that are used in these notes.")
kbl(tibble(read_csv("data/notation-abbrev.csv")), longtable = T, booktabs = T) %>%
 kable_styling(latex_options = c("striped", "repeat_header"))
```

<!--chapter:end:00-prelim.Rmd-->


# Sampling distributions {#sampling-distributions}

Placeholder


## Normal distribution {#normal-distribution}
### Some useful facts about Normals {#facts-normals}
### Empirical rule ($68-95-99.7$ rule)
## $\mathsf{t}$ distribution {#t-distribution}
### Properties of $\mathsf{t}$ distributions {#facts-t}
## $\chi^2$ distribution {#chisq-distribution}
## $\mathsf{F}$ distribution {#F-distribution}

<!--chapter:end:01-sampling-distributions.Rmd-->


# Basics of statistical inference {#statistical-inference}

Placeholder


## Point estimation {#point-estimation}
## Confidence intervals {#confidence-intervals}
## Hypothesis testing {#hypothesis-testing}

<!--chapter:end:02-basics-stat-infer.Rmd-->


# Inferences based on a single sample {#inference-single-sample}

Placeholder


## Estimating means {#estimating-means}
### Mean of a normal population with known variance {#mean-normal-var-known}
### Mean of a population with unknown variance (large-sample) {#mean-large-sample}
### Mean of a normal population with unknown variance {#mean-normal-var-unknown}
## Estimating proportions {#estimating-proportions}
## Estimating variances {#estimating-variances}

<!--chapter:end:03-infer-single-sample.Rmd-->


# Inferences based on two samples {#inference-two-samples}

Placeholder


## Comparing means {#compare-means}
### Comparing means of normal populations when variances are known {#compare-means-normpops-vars-known}
### Comparing means when the sample sizes are large {#compare-means-large-samples}
### Comparing means of normal populations when variances are unknown and the sample size is small {#compare-means-normpops-vars-unknown}
## Comparing paired samples {#compare-paired-samples}
## Comparing proportions {#compare-proportions}
## Comparing variances {#compare-variances}

<!--chapter:end:04-infer-two-samples.Rmd-->


# Analysis of variance (ANOVA) {#anova}

Placeholder


## Single factor ANOVA test {#anova-single-factor-test}
## Confidence intervals {#anova-ci}

<!--chapter:end:05-anova.Rmd-->


# Linear regression {#linear-models}

Placeholder


## Simple linear regression models {#simple-linear-regression}
## Estimating $\sigma^2$ for linear regressions {#ls-estimate-var}
## Inferences for least-squares parameters {#inference-ls}
## Correlation {#correlation}

<!--chapter:end:06-linear-models.Rmd-->


# Categorical data {#categorical-data}

Placeholder


## Multinomial experiments
## Goodness-of-fit for a single factor {#goodness-of-fit-tests}
## Test for independence of factors

<!--chapter:end:07-categorical-data.Rmd-->


# Quality Control {#quality-control}

Placeholder



<!--chapter:end:08-quality-control.Rmd-->

# (PART\*) Appendix {-}

# Curated Content {-}

Below we provide links to supplementary online material. Hopefully some of the items will inspire you to further examine different aspects of the course. 

## Investigation 0 {-}

What is Statistics? 

Cambridge Ideas - Professor Risk ::
[https://www.youtube.com/watch?v=a1PtQ67urG4](https://www.youtube.com/watch?v=a1PtQ67urG4) Prof David Spiegelhalter (Cambridge University) discusses public understanding of risk.

The Joy of Statistics ::
[https://www.youtube.com/watch?v=jbkSRLYSojo](https://www.youtube.com/watch?v=jbkSRLYSojo) Prof Hans Rosling (Karolinska Institute and Gapminder Foundation) analyzes data from 200 Countries over 200 Years in 4 Minutes - The Joy of Stats - BBC Four.

Teach statistics before calculus! ::
[https://www.ted.com/talks/arthur_benjamin_teach_statistics_before_calculus](https://www.ted.com/talks/arthur_benjamin_teach_statistics_before_calculus) Prof Arthur Benjamin (Harvey Mudd College) argues that the pinnacle of math education be probability and statistics --- not calculus.

## Investigation 1 {-}

Bad statistics / Good statistics

Three ways to spot bad statistics ::
[https://www.ted.com/talks/mona_chalabi_3_ways_to_spot_a_bad_statistic](https://www.ted.com/talks/mona_chalabi_3_ways_to_spot_a_bad_statistic) Mona Chalabi (Data Journalist) discusses three ways to spot bad statistics.

No. 10 daily Covid-19 briefings ::
[https://www.youtube.com/watch?v=QIw2l-trRXc](https://www.youtube.com/watch?v=QIw2l-trRXc) Prof David Spiegelhalter (Cambridge University) claims that No. 10 is misrepresenting Covid-19 data in a discussion with Andrew Marr (BBC).




## Investigation 2 {-}

Confounders.

How statistics can be misleading (TED-Ed) ::
[https://www.ted.com/talks/mark_liddell_how_statistics_can_be_misleading](https://www.ted.com/talks/mark_liddell_how_statistics_can_be_misleading) Mark Liddell (Educator) discusses Simpson's Paradox in this TED-Ed animation.

## Investigation 3 {-}


## Investigation 4 {-}


## Investigation 5 {-}


## Investigation 6 {-}


## Investigation 7 {-}


## Investigation 8 {-}


# References {-}

<!--chapter:end:10-investigations.Rmd-->

