---
title: "MA22004"
subtitle: "Seminar 10"
author: "Dr Eric Hall"
date: "10/12/2020"
output: 
 ioslides_presentation:
   widescreen: true
   css: ../assets/style.css
   logo: ../assets/images/rev_logo.png
---
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\E}{\mathbf{E}}
\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\corr}{corr}
\newcommand{\se}{\mathsf{se}}
\DeclareMathOperator{\sd}{sd}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = "", fig.asp = .5)
library(tidyverse)
library(latex2exp)
library(knitr)
library(kableExtra)
library(janitor)
beer <- read_csv("../../data/beer.csv")
ipa_new <- read_csv("../../data/beer2.csv")
ipa <- beer %>% 
 select(Batch_Id, pH, ABV, OG, IBU, Beer) %>% 
 filter(Beer == "IPA") %>% 
 rename(Day = Batch_Id)
ipa$Day[1:48] <- rep(1:16, each = 3)
ipa <- ipa[1:48,]
m <- 3    # three batches per day
k <- 16   # number of days

theme_ur <- theme(legend.justification = c(1,1), legend.position = c(1,1), legend.box.margin = margin(c(4, 4, 4, 4), unit = "pt"))
theme_lr <- theme(legend.justification = c(1,0), legend.position = c(1,0), legend.box.margin = margin(c(4, 4, 4, 4), unit = "pt"))
options(width = 90, knitr.kable.NA = '', "kableExtra.html.bsTable" = T)
lsz <- 1.0
tsz <- 4
```

## Announcements

### Reminders 

- It is week 10! This is our **last seminar before** the Class Test 3. 
- You should have read ยง8 Quality Control on **Perusall**.
- Please look over the feedback from Class Test 2. 

### Upcoming
- Lab 7 due **Thursday 10 Dec** at **17:00** (late accepted until Sunday w/o penalty).
- Worksheet #10 Review is already posted to Blackboard.
- Class Test 3 is **Wednesday 16 Dec** from **14:00--17:00**.


## What will Class Test 3 cover?

Class Test 3 will be longer than the first two tests and will be cumulative in nature, but will focus on the following topics.

- Categorical data ($\chi^2$ tests for goodness of fit and independence)
- Linear regression (linear models, least squares estimates, inference for variances and least squares parameters, correlation)
- ANOVA (single factor tests and confidence intervals)

:::{.warningblock}
Revise inferences (hypothesis tests and confidence intervals) for single samples and two samples. 
:::

Monday's workshop will include a review of important topics. 


## Quality Control (QC)

- Identify the sources of random variations in output processes that might have *assignable causes*.
- **Control charts** help us to recognize when industrial processes are no longer controlled.
- Control charts use time series data to calculate the running value of a quality statistic. 
- If the quality statistic exceed upper/lower control limits, process is deemed to be out of control (quality negatively impacted).

:::{.noteblock}
Workflow:

time series $\longrightarrow$ control chart $\longrightarrow$ identify assignable causes 
:::

## Beer Data: IPA

A brewer produces 3 batches of IPA per day and measures the pH value of each batch, which influences saccharification. 

```{r scatter, echo = TRUE, eval = TRUE}
ggplot(ipa, aes(x = Day, y = pH)) + geom_point(size = 2)
```


## $3\sigma$ Control Chart

We consider the typical $3\sigma$ control chart for the mean pH $\overline{X}$.

Assume generating process $X$ is normally distributed with unknown $\mu$ and $\sigma$. 

Our control region is specified to by three standard deviations: the process remains in control if it stays three deviations within a baseline value. 

:::{.tipblock}
Why consider three standard deviations from the mean?
:::

## Checking for normality: `qqplot`

```{r qqplot, echo = TRUE, eval = TRUE}
ggplot(ipa, aes(sample = pH)) + stat_qq(size = 2) + stat_qq_line(color = "purple")
```

## Checking for normality: `shapiro.test`

Shapiro--Wilk tests the null hypothesis that a given sample came from a normally distributed population. Thus if $P$-value small, reject null and do no proceed. 

```{r shapiro, echo = TRUE, eval = TRUE}
shapiro.test(ipa$pH)
```

:::{.warningblock}
Failing to reject null does *not* tell us the population is normally distributed...
:::

## Checking for normality ?

:::{.tipblock}
If the sample size is *large*, which should we use (`qqplot` vs `shapiro.test`)?
:::

## Constructing a control chart

```{r ipa-stat, echo = FALSE, eval = TRUE, message = FALSE}
ipa_stat <- ipa %>% 
   group_by(Day) %>% 
   summarise(observations = list(pH), mean = signif(mean(pH), digits = 4), 
             sd = signif(sd(pH), digits = 4), range = max(pH) - min(pH))
   
kbl(head(ipa_stat), align = "rlccc", 
    col.names = c('Day', "pH Observations", '$\\overline{x}$', '$s$', 'Range'),
    booktabs = T) %>%
 kable_styling(latex_options = c("hold_position")) 
```

$\widehat{\mu}$ 


$\widehat{\sigma}$

## Estimating $\sigma$

Recall that the sample standard deviation is a biased estimator for the standard deviation. 

For $X_1, \dots, X_n \sim \mathsf{N}(\mu,\sigma^2)$,
\[\E(S) = a_n \cdot \sigma\]
where the *bias correction* can be computed exactly:
\[a_n = \sqrt{\frac{2}{n-1}}\frac{\Gamma(\tfrac{n}{2})}{\Gamma(\tfrac{n-1}{2})} \,.\]

```{r bias-correction, echo = FALSE, eval = TRUE, message = FALSE}
cor <- rbind(c("$n$",seq(3,8,1)), c("$a_n$", .886,.921,.940,.952,.959,.965))
kbl(cor, booktabs = T) %>%
 kable_styling(latex_options = c("hold_position")) 
```

## Unbiased estimator $\sigma$

Let $\overline{S} = \frac{1}{k} \sum_{i=1}^k S_i$, where $S_i$ are the sample sd for $k$ days.

<!-- Then  -->
<!-- \[\E(\overline{S}) = \frac{1}{k} \E\left(\sum_{i=1}^k S_i\right) = -->
<!-- \frac{1}{k} \sum_{i=1}^{k} \E(S_i) = \frac{1}{k} \sum_{i=1}^{k} a_n \sigma  = an \sigma \] -->

<!-- Thus -->
<!-- \[\E(\overline{S}/a_n) = \frac{1}{a_n} \E(\overline{S}) = \frac{1}{a_n} a_n \sigma = \sigma \] -->

<!-- so $\widehat{\sigma} = \overline{S} / a_n$ is an unbiased estimator of $\sigma$.  -->


## Control limits based on sample sds

\[\mathsf{LCL} = \widehat{\mu} - 3 \frac{\overline{s}}{a_n \sqrt{n}}\]
\[\mathsf{UCL} = \widehat{\mu} + 3 \frac{\overline{s}}{a_n \sqrt{n}}\]
where 

$\widehat{\mu} = \frac{1}{k} \sum_{i=1}^k \overline{x}_i$

$\overline{s} = \frac{1}{k} \sum_{i=1}^k s_i$

## Control Chart

- Assume process is in-control over the 16 days and compute the limits. 
- Violation would require that we seek to identify an assignable cause.

```{r qc-beer-control-chart, echo = FALSE, warning = FALSE, message = FALSE}
a <- function(m){ sqrt(2) * gamma(m/2) / (sqrt(m-1) * gamma((m-1)/2)) }
muhat = sum(ipa_stat$mean) / k
sbar = sum(ipa_stat$sd) / k
lcl = muhat - 3*sbar / (a(m) * sqrt(m))
ucl = muhat + 3*sbar / (a(m) * sqrt(m))

ggplot(ipa_new, aes(x = Day)) + geom_point(aes(y = mean)) + 
 geom_hline(aes(yintercept = muhat, color = "Mean"), size = lsz) + 
 geom_hline(aes(yintercept = lcl, color = "LCL"), size = lsz*1.5) + 
 geom_hline(aes(yintercept = ucl, color = "UCL"), size = lsz*1.5) + ylab("pH") + 
   theme(legend.justification = c(1,1), legend.position = c(1,1),
         legend.title = element_blank(), 
         legend.box.margin = margin(c(4, 4, 4, 4), unit = "pt"))
```  

## Summary

Today we discussed control charts for Quality Control. 

We learned about bias correction for estimating the population standard deviation. 
